---
title: "Analysis on Historical Crime Data for New York City"
author: "Molly pitts"
date: "11/18/2021"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
pacman::p_load(devtools, tidyverse, dplyr, nycmaps, maps, usmap, cowplot, lubridate, RColorBrewer, gridExtra, tigris, sp, ggmap, maptools, broom, httr, rgdal, maps, wesanderson, ggplot2, kableExtra, DT, htmltools, htmlwidgets, gtools, rvest, tidytext, xml2, rio, here, flexdashboard, shiny, plotly)
# devtools::install_github("zachcp/nycmaps")
```

```{r borough-map-data, message=FALSE, warning = FALSE}
# nyc BOROUGHS mapping info

# This code chunk imports a dataset that contains mapping data for the boroughs of New York City. I wrangled
# the dataset by renaming all of the regions included in the dataset to just the name of the borough they are located in,
# so joining with future datasets would be simpler.


# map(database="nyc")
nyc <- map_data("nyc")
nyc <- nyc %>%
  mutate(Borough = case_when(grepl("Manhattan", region) ~ "Manhattan",
                             grepl("Bronx", region) ~"Bronx",
                             grepl("Brook", region) ~ "Brooklyn",
                             grepl("Queens", region) ~ "Queens",
                             grepl("Staten Island", region) ~ "Staten Island"))
nyc$subregion <- NULL

# This is an example of the graph that this dataset produces.
# nyc_borough_plot <- nyc %>%
#   ggplot() + 
#   geom_map(map=nyc,
#            color = "gray",
#            mapping = aes(x=long, 
#                y=lat, 
#                map_id=region)) +
#   coord_equal() +
#   theme_map()
```

```{r wrangling-historical-crime-1, message=FALSE, warning = FALSE}
# This code chunk wrangles historical felony data from 2000 to 2020 in New York City.
# importing raw historical felony data from csv
hist_felony_raw <- read_csv("data/seven-major-felony-offenses-2000-2020.csv")
hist_felony_data <- hist_felony_raw

# wrangling of first dataset. variables names are changed to make data more concise. data is pivoted longer
# to allow data to be sorted by year. The row that provides the total sum of felonies is removed as it is not needed.
hist_felony_data <- hist_felony_data  %>%
  rename(Felony = OFFENSE) %>%
  mutate(Felony = str_replace(Felony,
                              "MURDER & NON-NEGL. MANSLAUGHTER",
                              "MURDER"),
         Felony = str_replace(Felony,
                              "FELONY ASSAULT",
                              "ASSAULT"),
         Felony = str_replace(Felony,
                              "GRAND LARCENY OF MOTOR VEHICLE",
                              "GRAND LARCENY OF VEHICLE")) %>%
  pivot_longer(cols = `2000`:`2020`, 
               names_to = "Year", 
               values_to = "Number") %>%
  filter(Felony != "TOTAL SEVEN MAJOR FELONY OFFENSES")

# This is an html widget that allows the viewing and searching of this datatable, as it is one of the most important
# datasets in this project.
datatable(hist_felony_data, options = list(pageLength = 5))

```

```{r wrangling-historical-pop-1, message=FALSE, warning = FALSE}
# This code chunk wrangles historical population data for New York City.
# importing raw historical population data from csv
hist_borough_pop_raw <- read_csv("data/new-york-city-population-by-borough-1950-2040.csv")
hist_borough_pop <- hist_borough_pop_raw 

# wrangling of second dataset. the populations for the years 2000, 2010, and 2020 were selected. The total
# New York City population was selected for each year. The dataset was pivoted longer to make the data
# more usable. The individual borough populations were removed.
hist_borough_pop_tot <- hist_borough_pop %>%
  select(`Borough`,`2000`,`2010`,`2020`) %>%
  filter(Borough == "NYC Total") %>%
    pivot_longer(cols = `2000`:`2020`, 
               names_to = "Year", 
               values_to = "NycPop")
hist_borough_pop_tot$Borough = NULL

# the same raw population dataset is wrangled to display the populations for each borough. 
hist_borough_pop <- hist_borough_pop %>%
  select(`Borough`,`2000`,`2010`,`2020`) %>%
  filter(Borough != "NYC Total") 
hist_borough_pop_longer <- hist_borough_pop %>%
  pivot_longer(cols = `2000`:`2020`, 
               names_to = "Year", 
               values_to = "Pop")

# data for the year 2000 is selected from the dataset
nyc_pop_2000 <- hist_borough_pop %>%
  select(`Borough`,`2000`) %>%
  rename(Pop = `2000`)

# data for the year 2010 is selected from the dataset
nyc_pop_2010 <- hist_borough_pop %>%
  select(`Borough`,`2010`) %>%
  rename(Pop = `2010`)

# data for the year 2020 is selected from the dataset
nyc_pop_2020 <- hist_borough_pop %>%
  select(`Borough`,`2020`) %>%
  rename(Pop = `2020`)
```

```{r line-graph-1, message=FALSE, warning = FALSE}
# This code chunk generates a line graph that displays the evolution of the numbers of felonies committed from 2000 to 2020
# in New York City. This is total felonies, not per capita.

# Line graph showing all felonies in nyc from 2000-2020
colors = c(wes_palette("GrandBudapest1"), wes_palette("GrandBudapest2"))
hist_felony_plot_2 <- hist_felony_data %>%
  ggplot(aes(x = Year, y = Number, group = Felony, color = Felony)) +
  geom_line(size = 1) +
  labs(x = NULL, y = "Number of Cases", title = "Number of Felonies Committed", subtitle = "Over The Past Twenty Years in NYC") +
  scale_color_manual(values = colors) +
  theme(legend.position = "top", 
        axis.text.x = element_text(angle = 90),
        legend.title = element_blank())
hist_felony_plot_2
```

```{r wrangling-historical-crime-2, message=FALSE, warning = FALSE}
# This code chunk wrangles historical felony data by precinct, which allows the historical felony dataset to be connected to the New York City Borough mapping data.

# importing raw historical felony by precinct data from csv
hist_felony_by_precinct_raw <- read_csv("data/seven-major-felony-offenses-by-precinct-2000-2020.csv")
hist_felony_by_precinct_data <- hist_felony_by_precinct_raw 

# wrangling fourth dataset.
# dataset wrangled with FOUR pipes.
# column names are renamed for clarity, felony names are renamed for readability and simplicity, the dataset is pivoted longer to allow for sorting by year, total sum of felonies is removed because it is not needed, precinct numbers are used to assign rows to boroughs, the precinct column is removed because it is no longer needed, and the dataset is finally grouped by borough and felony to facilitate the next graphs.
hist_felony_by_precinct_data <- hist_felony_by_precinct_data %>%
  rename(Felony = CRIME,
         Precinct = PCT) %>%
  mutate(Felony = str_replace(Felony,
                              "MURDER & NON NEGL. MANSLAUGHTER",
                              "MURDER"),
         Felony = str_replace(Felony,
                              "FELONY ASSAULT",
                              "ASSAULT"),
         Felony = str_replace(Felony,
                              "GRAND LARCENY OF MOTOR VEHICLE",
                              "GRAND LARCENY OF VEHICLE")) %>%
  pivot_longer(cols = `2000`:`2020`, 
               names_to = "Year", 
               values_to = "Number") %>%
  filter(Felony != "TOTAL SEVEN MAJOR FELONY OFFENSES") %>%
  mutate(Borough = case_when((Precinct < 35) ~ "Manhattan",
                             (Precinct > 34 & Precinct < 53) ~ "Bronx",
                             (Precinct > 52 & Precinct < 94) ~ "Brooklyn",
                             (Precinct > 93 & Precinct < 120) ~ "Queens",
                             (Precinct > 119) ~ "Staten Island")) %>%
  select(-Precinct) %>%
  group_by(Borough, Felony)

# shows the total felonies of each type committed in each borough from 2000-2020. gets rid of na values in the number column.
hist_felony_by_precinct_data_tot <- hist_felony_by_precinct_data
hist_felony_by_precinct_data_tot <- hist_felony_by_precinct_data_tot %>%
  filter(!is.na(Number)) %>%
  ungroup() %>%
  group_by(Borough, Felony, Year) %>%
  mutate(Total = sum(Number)) %>%
  select(-Number) %>%
  distinct()
```

```{r wrangling-crime-per-capita, messafe=FALSE, warning=FALSE}
# This code chunk wrangles data to allow for the generation of graphs that display information regarding the total number of felonies committed in each borough per capita for the years 2000, 2010, and 2020.

# wrangling data to graph felonies committed by borough for 2000
felonies_boroughs_2000 <- hist_felony_by_precinct_data 
felonies_boroughs_2000 <- felonies_boroughs_2000 %>%
  filter(Year == "2000" & !is.na(Number)) %>%
  mutate(Total = sum(Number)) %>%
  select(-Number) %>%
  distinct() 

# shows the top felonies committed in each borough in 2000
felonies_boroughs_2000_top <- felonies_boroughs_2000 
felonies_boroughs_2000_top <- felonies_boroughs_2000_top %>%
  group_by(Borough) %>%
  top_n(1, Total)

# adds mapping data and population data
felonies_boroughs_2000_p <- felonies_boroughs_2000 %>%
  left_join(nyc_pop_2000, by = "Borough") %>%
  mutate(PerCapita = Total/Pop) %>%
  left_join(nyc, by = "Borough")

# ---

# wrangling data to graph felonies committed by borough for 2010
felonies_boroughs_2010 <- hist_felony_by_precinct_data 
felonies_boroughs_2010 <- felonies_boroughs_2010 %>%
  filter(Year == "2010" & !is.na(Number)) %>%
  mutate(Total = sum(Number)) %>%
  select(-Number) %>%
  distinct()

# shows the top felonies committed in each borough in 2010
felonies_boroughs_2010_top <- felonies_boroughs_2010 
felonies_boroughs_2010_top <- felonies_boroughs_2010_top %>%
  group_by(Borough) %>%
  top_n(1, Total)

# adds mapping data and population data
felonies_boroughs_2010_p <- felonies_boroughs_2010 %>%
  left_join(nyc_pop_2010, by = "Borough") %>%
  mutate(PerCapita = Total/Pop) %>%
  left_join(nyc, by = "Borough")

# ---

# wrangling data to graph felonies committed by borough for 2020
felonies_boroughs_2020 <- hist_felony_by_precinct_data 
felonies_boroughs_2020 <- felonies_boroughs_2020 %>%
  filter(Year == "2020" & !is.na(Number)) %>%
  mutate(Total = sum(Number)) %>%
  select(-Number) %>%
  distinct() 

# shows the top felonies committed in each borough in 2020
felonies_boroughs_2020_top <- felonies_boroughs_2020 
felonies_boroughs_2020_top <- felonies_boroughs_2020_top %>%
  group_by(Borough) %>%
  top_n(1, Total)

# adds mapping data and population data
felonies_boroughs_2020_p <- felonies_boroughs_2020 %>%
  left_join(nyc_pop_2020, by = "Borough") %>%
  mutate(PerCapita = Total/Pop) %>%
  left_join(nyc, by = "Borough")
```

```{r three-decade-borough-maps, message=FALSE, warning = FALSE}
# This code chunk generates three plots. Each plot shows the numbers of total felonies committed per capita in each borough of New York City in the years 2000, 2010, and 2020.
# Map of Number of Felonies Per Capita committed in each Borough in 2000.
felonies_boroughs_2000_plot <- felonies_boroughs_2000_p %>%
  mutate(Felony = str_replace(Felony,
                              "OF VEHICLE",
                              "AUTO")) %>%
  ggplot() +
  geom_map(map = nyc,
           color = "gray",
           mapping = aes(x = long,
               y = lat,
               map_id = region,
               fill = PerCapita)) +
  coord_equal() +
  theme_map() +
  theme(legend.position = "bottom",
        legend.key.width = unit(1, 'cm')) +
  scale_fill_distiller(palette="Blues") +
  facet_wrap(~Felony, labeller = labeller(group = label_wrap_gen(width = 50))) +
  labs(title = "Total Felonies Per Capita", subtitle = "in 2000")

# Map of Number of Felonies committed in each Borough in 2010.
felonies_boroughs_2010_plot <- felonies_boroughs_2010_p %>%
  mutate(Felony = str_replace(Felony,
                              "OF VEHICLE",
                              "AUTO")) %>%
  ggplot() +
  geom_map(map = nyc,
           color = "gray",
           mapping = aes(x = long,
               y = lat,
               map_id = region,
               fill = PerCapita)) +
  coord_equal() +
  theme_map() +
  theme(legend.position = "bottom",
        legend.key.width = unit(1, 'cm')) +
  scale_fill_distiller(palette="Blues") +
  facet_wrap(~Felony) +
  labs(title = "Total Felonies Per Capita", subtitle = "in 2010")

# Map of Number of Felonies committed in each Borough in 2020.
felonies_boroughs_2020_plot <- felonies_boroughs_2020_p %>%
  mutate(Felony = str_replace(Felony,
                              "OF VEHICLE",
                              "AUTO")) %>%
  ggplot() +
  geom_map(map = nyc,
           color = "gray",
           mapping = aes(x = long,
               y = lat,
               map_id = region,
               fill = PerCapita)) +
  coord_equal() +
  theme_map() +
  theme(legend.position = "bottom",
        legend.key.width = unit(1, 'cm')) +
  scale_fill_distiller(palette="Blues") +
  facet_wrap(~Felony) +
  labs(title = "Total Felonies Per Capita", subtitle = "in 2020")

felonies_boroughs_2000_plot
felonies_boroughs_2010_plot
felonies_boroughs_2020_plot
```


```{r three-decade-bar-charts, message=FALSE, warning = FALSE}
# Three bar plots for felonies committed in nyc in 2000, 2010, and 2020.
colors = c(wes_palette("Darjeeling1"), wes_palette("Royal1"))
hist_felony_plot_3 <- hist_felony_data %>%
  filter(Year == "2000" |
         Year == "2010" |
         Year == "2020") %>%
  arrange(desc(Number)) %>%
  ggplot(aes(x = Felony,
           y = Number,
           fill = Felony)) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "Number", title = "Felonies Committed in NYC in 2000, 2010, and 2020", fill = "") +
  scale_fill_manual(values = colors) +
  theme_minimal() +
  theme(legend.position = "top",
        plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank()) +
  facet_wrap(~Year)
hist_felony_plot_3

colors2 = c(wes_palette("GrandBudapest1"), wes_palette("GrandBudapest2"))
hist_felony_plot_4 <- hist_felony_data %>%
  filter(Year == "2000" |
         Year == "2010" |
         Year == "2020") %>%
  arrange(desc(Number)) %>%
  left_join(hist_borough_pop_tot, by="Year") %>%
  mutate(PerCap = Number/NycPop) %>%
  ggplot(aes(x = Felony,
           y = PerCap,
           fill = Felony)) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "Felonies Per Capita", title = "Felonies Committed Per Capita in NYC in 2000, 2010, and 2020", fill = "") +
  scale_fill_manual(values = colors2) +
  theme_minimal() +
  theme(legend.position = "top",
        plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank()) +
  facet_wrap(~Year)
hist_felony_plot_4
```

<br>
<br>
<br>
<br>
<br>

### Analysis

1. What were the most common felonies committed in the city of New York in the years 2000, 2010, and 2020?<br>
  + The most common felonies committed in New York City by total amount and per capita were Grand Larceny in all three years. As seen in the graphs, Grand Larceny is far more common than other crimes. In New York City, Grand larceny involves the theft of property values at over $1000. It also applies to theft of any firearm or vehicle, and can apply to extortion and blackmail. ADD WHY))))))))))))))))<br>
<br>

```{r answer-1, echo=FALSE}
hist_felony_plot_3
hist_felony_plot_4
```

<br>
2. What were the most common felonies committed in each borough of New York in 2000, 2010, and 2020?<br>
  + The most common felonies committed in each borough are displayed in the tables below:
  In the year 2000 there were a variety of felonies being committed most often. As time goes on, the trends change. As seen across this report, as overall crime has decreased in New York, Grand Larceny has decreased the slowest, which has made it the most commonly committed felony in New York City today. ADD WHY))))))))))))))))
<br>

```{r answer-2, echo=FALSE}
# using kableExtra tables to display top felonies
felonies_boroughs_2000_top <- felonies_boroughs_2000_top[,c(2,3,1,4)]
felonies_boroughs_2000_top %>%
  rename(`Most Common Felony` = "Felony") %>%
  kbl(caption = "Top felonies for each borough of NYC in 2000") %>%
  kable_classic(full_width = F, html_font = "Cambria", position = "center")

felonies_boroughs_2010_top <- felonies_boroughs_2010_top[,c(2,3,1,4)]
felonies_boroughs_2010_top %>%
  rename(`Most Common Felony` = "Felony") %>%
  kbl(caption = "Top felonies for each borough of NYC in 2010") %>%
  kable_classic(full_width = F, html_font = "Cambria", position = "center")

felonies_boroughs_2020_top <- felonies_boroughs_2020_top[,c(2,3,1,4)]
felonies_boroughs_2020_top %>%
  rename(`Most Common Felony` = "Felony") %>%
  kbl(caption = "Top felonies for each borough of NYC in 2020") %>%
  kable_classic(full_width = F, html_font = "Cambria", position = "center")
```

<br>
3. How have felony rates changed over the past two decades?<br>
  + Overall crime in New York has decreased. In 2019, burglary and grand larceny of vehicles increased, while general grand larceny decreased. Also, there was a period of time around 2010 where grand larceny had decreased significantly. Overall, however, Grand Larceny has become the most commonly committed crime in New York City. ADD WHY))))))))))))))))) <br><br>
<br> 

```{r answer-3, echo=FALSE}
hist_felony_plot_2
```

```{r answer-3-a, echo=FALSE}
felonies_boroughs_2000_plot
felonies_boroughs_2010_plot
felonies_boroughs_2020_plot
```

# posted 

### posted  Questions

1. What boroughs within NY have the highest/lowest household income?<br>
	Manhattan anchored by Wall Street shows as one of the world's premiere financial centers. As well as housing the NASDAQ and NYSE stock exchanges. This shows in their wealth.

	Brooklyn, Staten Island, and Queens almost have a bell curve minus a high amount of 100K-199K income levels

	Bronx has an intensely disproportionate income levels versus the rest of the boroughs and was continuously ranked last among the boroughs.
<br><br>
2. In what ways have the household income levels improved or worsened over the past 15 years?<br>
	Bronx: Has steadily remained the borough with the highest population of poor families, with a small decline the last few years.

	Brooklyn: Has “improved” as throughout the years more people with >100K have populated the area.

	Manhattan: Throughout the years it’s become populated with more wealth as anything below 100K income slowly fades.

	Staten Island: Remains about half rich as of today, with the poorer steadily decreasing.
<br><br>
3. Is there a correlation between low income districts and crime rates?<br>
	Manhattan had the largest decline in felonies over the years

	Staten Island has kept a low rate of felonies over its years even despite its smaller population

	I believe in Manhattan you are more likely to charged with a felony because more people can afford to take you to court.

	Brooklyn has the most felonies among the more sizeable populations

Bronx has only slightly decreased in felonies
<br><br>

```{r pp-wrangling}

#Importing my datasets and assigning them
family_Income_raw <- read.csv("data/Family Income.csv", stringsAsFactors = T)
#median_Income_raw <- read_csv("data/Median Incomes.csv")

family_Income_data <- family_Income_raw
#median_Income_data <- median_Income_raw

#Create a list to sort the data by the named Boroughs
borough <- c("Brooklyn", "Queens" , "Staten Island", "Manhattan", "Bronx")


#where the filtering takes place
family_Income_data <- family_Income_data %>% 
  filter(Location %in% borough & DataFormat == "Number") %>% 
  rename(Boroughs = Location) %>% 
  rename(Number = Data) %>% 
  select(-DataFormat, -Fips)


#I performed this operation to help me properly sort my graphs as I could not get the boroughs to sort from highest income to lowest
#So I sorted the strings into numbers to sort by
family_Income_data <- mutate(family_Income_data, Level = ifelse(Income.Level == 	"Under $15,000", 1,
                                                                ifelse(Income.Level == "$15,000 to $24,999", 2,
                                                                ifelse(Income.Level == "$25,000 to $34,999", 3,
                                                                ifelse(Income.Level == "$35,000 to $49,999", 4,
                                                                ifelse(Income.Level == "$50,000 to $74,999", 5,
                                                                ifelse(Income.Level == "$75,000 to $99,999", 6,
                                                                ifelse(Income.Level == "$100,000 to $199,999", 7,
                                                                ifelse(Income.Level == "$200,000 or more", 8,0
                                                                              )))))))))

family_Income_data <- family_Income_data[order(family_Income_data$Boroughs, family_Income_data$TimeFrame, family_Income_data$Level),]

rownames(family_Income_data) = seq(length = nrow(family_Income_data))

family_Income_data$Income.Level <- factor(family_Income_data$Income.Level, levels = c("$200,000 or more",
                                 "$100,000 to $199,999",
                                 "$75,000 to $99,999",
                                 "$50,000 to $74,999",
                                 "$35,000 to $49,999",
                                 "$25,000 to $34,999",
                                 "$15,000 to $24,999",
                                 "Under $15,000"))

```

```{r posted, message=FALSE, warning = FALSE}

#This keeps my numbers from showing in scientific notation
options(scipen = 999)

#Making my fun multilayed plot, this will group together all 5 boroughs, along side all 8 income levels
#along side our 14 years. Giving us 5 * 8 * 14 sections to look at
family_Income_Plot <- family_Income_data %>%
  ggplot(mapping = aes(x = TimeFrame,
             y = Number,
             fill = factor(Income.Level))) +
  geom_bar(size = 1, stat = "identity") +
  labs(x = NULL,
       y = "Number of People",
       title = "Population Income Grouping from 2005 - 2019") +
  theme(legend.position = "left",
        axis.text.x = element_text(angle = 90,
                                   size = 5),
        legend.title = element_blank(),
        plot.title = element_text(size = 15))+
  facet_wrap(~Boroughs, scales = "free")
  

family_Income_Plot
```

```{r posted2, message=FALSE, warning = FALSE}

#I wanted a clean view of 15 years summarized so we can see general conditions over time
fifteen_Year_Avg <- family_Income_data

fifteen_Year_Avg <- fifteen_Year_Avg %>% 
  group_by(Boroughs, Income.Level) %>% 
  summarize(income_avg = mean(Number))

#Reordering my levels on the legend, not having them from greatest to least really bothered me
fifteen_Year_Avg$Income.Level <- factor(fifteen_Year_Avg$Income.Level, levels = c("$200,000 or more",
                                 "$100,000 to $199,999",
                                 "$75,000 to $99,999",
                                 "$50,000 to $74,999",
                                 "$35,000 to $49,999",
                                 "$25,000 to $34,999",
                                 "$15,000 to $24,999",
                                 "Under $15,000"))

#Making the bar plot that limits the amount of years down to just the borough average's
fifteen_Year_Avg_Plot <- fifteen_Year_Avg %>% 
  ggplot(mapping = aes(x = Income.Level,
             y = income_avg,
             fill = factor(Income.Level),
             shape = Income.Level)) +
  geom_bar(size = 1, stat = "identity") +
  labs(x = NULL,
       y = "Number of People",
       title = "Population Income Grouping Average") +
  theme(legend.position = "left",
        axis.text.x = element_blank(),
        legend.title = element_blank(),
        plot.title = element_text(size = 15))+
  facet_wrap(~Boroughs, scales = "free")

fifteen_Year_Avg_Plot

```

```{r posted3, message=FALSE, warning = FALSE}

#limit Molly's data to the years I have in mine 2005 to 2019
felony_tot <- subset(hist_felony_by_precinct_data_tot, Year == "2005" | Year == "2006" | Year == "2007" | Year == "2008" | Year == "2009" | Year == "2010" | Year == "2011" | Year == "2012" | Year == "2013" | Year == "2014" | Year == "2015" | Year == "2016" | Year == "2017" | Year == "2018" | Year == "2019")

#Rid the na values
felony_tot <- na.omit(felony_tot)

#Total and match all the years with the total crimes in that borough
felony_tot <- felony_tot %>% 
  group_by(Year, Borough) %>% 
  mutate(crime_Total = sum(Total)) %>% 
  select(-Felony) %>%
  rename(Boroughs = Borough) %>% 
  distinct(crime_Total)

felony_tot <- na.omit(felony_tot)

family_Income_data <- family_Income_data %>% 
  select(-Level)

crime_Income_Relation <- merge(family_Income_data, felony_tot, by.x = c("Boroughs", "TimeFrame"), by.y = c("Boroughs", "Year"))


#MAKE A BAR AND LINE GRAPH, plotting the line graph atop the income levels to help us see trends
crime_Income_Relation_Plot <- crime_Income_Relation %>% 
  ggplot(mapping = aes(x = TimeFrame,
             fill = factor(Income.Level))) +
  geom_bar(aes(y=Number), size = 1, stat = "identity") +
  geom_line(aes(y = 10*crime_Total), size = 1, color="red", group = 1, stat = "identity") +
  scale_y_continuous(sec.axis = sec_axis(~.*0.1, name = "Crimes Committed")) +
  labs(x = NULL,
       y = "Number of People",
       title = "Population Income Vs Crimes Committed that Year") +
  theme(legend.position = "left",
        axis.text.x = element_text(angle = 90,
                                   size = 5),
        legend.title = element_blank(),
        plot.title = element_text(size = 15))+
  facet_wrap(~Boroughs, scale = "free_y")

crime_Income_Relation_Plot

```

# trayed 

#Census data
```{r trayed-load-censusHTML, message=FALSE, warning = FALSE}
#todo: draw more demographic data from census
# page <- read_html(
#   "https://www.census.gov/quickfacts/fact/table/NY,newyorkcountynewyork,kingscountynewyork,queenscountynewyork,richmondcountynewyork,bronxcountynewyork/PST045219")
# page %>% xml2::xml_structure()
# 
# county_name <- page %>%
#   html_nodes(".qf-persist span") %>%
#   html_text() %>%
#   str_replace(", New York", "")
# 
# county_population <- page %>%
#   html_nodes(".icon-user-3~ tbody .selected td+ td") %>%
#   html_text() %>%
#   trimws() %>%
#   str_remove_all(",") %>%
#   as.numeric() %>%
#   glimpse()
# 
# poverty_level <- page %>%
#   html_nodes("tbody:nth-child(14) .fact~ .fact+ .fact td+ td") %>%
#   html_text() %>%
#   trimws() %>%
#   str_sub(4) %>%
#   str_remove("%") %>%
#   as.numeric() %>%
#   glimpse()
# 
# Borough_name <- c("New York Total", "Manhattan", "Brooklyn", "Queens", 
#                   "Staten Island", "Bronx")
# 
# NY_Census <- tibble(borough = Borough_name, county = county_name, 
#                     Population = county_population, 
#                     `poverty level` = poverty_level)
# 
# NY_Census %>%
#   write.csv("data/NYC_census_data.csv")

NY_census <- read.csv(file = "data/NYC_census_data.csv") %>%
  select(2:5)
```

#Temporary map (only reports population in borough)
```{r trayed-us-map-temp, message=FALSE, warning = FALSE}
us_states <- us_map("state")%>%
  rename(state = abbr)%>%
  rename(long = x)%>%
  rename(lat = y)

counties <- us_map("counties")

countiesNY <- counties %>%
  filter(abbr == "NY")

county_map <- left_join(countiesNY, NY_census, by = "county") %>%
  na.omit() %>%
  rename(long = x, lat = y)

county_map %>%
  ggplot(mapping = aes( x = long, y = lat, group = group, fill = Population))+
  geom_polygon()+
  scale_color_distiller(palette = "Spectral", aesthetics = "fill")+
  theme_map()+
  coord_equal()+
  labs(title = "Borough population")+
  theme_map()
```

#handgun permit (borough) data
```{r trayed-gun-data-boroughs, message=FALSE, warning = FALSE}
#https://www.nydailynews.com/new-york/ny-nypd-gun-ownership-data-20210517-jfodywqu25bvrijmcrtcaocpn4-story.html
gun_ownership_borough <- read_csv("data/handgun_permits.csv") %>%
  select(3:6) %>%
  rename(count = `Count by Zip code`)

gun_ownership_borough <- gun_ownership_borough %>%
  group_by(County) %>%
  mutate(sum = sum(count))

gun_ownership_borough$County <- gun_ownership_borough$County %>%
  trimws() %>%
  str_to_title() %>%
  paste("County", sep = " ")

gun_ownership_borough <- gun_ownership_borough %>%
  select(1,5) %>%
  rename(county = County)
```


```{r visualize-gun-data-Borough, message=FALSE, warning = FALSE}
gun_borough_map <- left_join(gun_ownership_borough, county_map,  by = "county")

gun_borough_map %>%
  ggplot(mapping = aes( x = long, y = lat, group = group, fill = sum))+
  geom_polygon()+
  scale_color_distiller(palette = "Spectral", aesthetics = "fill")+
  theme_map()+
  coord_equal()+
  labs(title = "Borough handgun Permits", fill = "Total Permits")+
  theme_map()
```

#gun ownership (state) data

```{r trayed-gun-data-states, message=FALSE, warning = FALSE}
gun_US <- read_csv("data/Gun_ownership_state.csv") %>%
  #https://worldpopulationreview.com/state-rankings/gun-ownership-by-state
  rename(full = State) %>%
  rename(sum = totalGuns) %>%
  select(1,3)

gun_US %>%
  rename(State = full,`Total Guns` = sum) %>%
  arrange(desc(`Total Guns`)) %>%
  datatable(options = list(pageLength = 5))
```


```{r visualize-gun-data-US, message=FALSE, warning = FALSE}
#Todo use much more detailed chloropleth map. ggplot for placeholding. 
gun_mapUS <- left_join(gun_US, us_states, by = "full")  
gun_mapUS %>%
  ggplot(mapping = aes( x = long, y = lat, group = group, fill = sum))+
  geom_polygon()+
  scale_color_distiller(palette = "Reds", aesthetics = "fill")+
  theme_map()+
  coord_equal()+
  labs(title = "Gun ownership total by State", fill = "Total Guns")+
  theme_map()
```

Sources:
NYC.gov: Historical New York City Crime Data: Citywide Seven Major Felony Offenses 2000-2020 & Historical New York City Crime Data: Citywide Seven Major Felony Offenses by Precinct 2000-2020 
https://www1.nyc.gov/site/nypd/stats/crime-statistics/historical.page
NYCOpen: New York City Population by Borough, 1950 - 2040 
https://data.cityofnewyork.us/City-Government/New-York-City-Population-by-Borough-1950-2040/xywu-7bv9
Github: nyc by zachcb 
https://github.com/zachcp/nycmaps
U.S. Census Bureau: Census data for NY boroughs 
https://www.census.gov/quickfacts/newyorkcitynewyork
