---
title: "Analysis on Historical Crime Data for New York City"
author: "Molly Pitts"
date: "2/18/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
pacman::p_load(devtools, tidyverse, dplyr, nycmaps, maps, usmap, cowplot, lubridate, RColorBrewer, gridExtra, tigris, sp, ggmap, maptools, broom, httr, rgdal, maps, wesanderson, ggplot2, kableExtra, DT, htmltools, htmlwidgets, gtools, rvest, tidytext, xml2, rio, here, flexdashboard, shiny, plotly)
# devtools::install_github("zachcp/nycmaps")
```

```{r borough-map-data, message=FALSE, warning = FALSE}
# nyc BOROUGHS mapping info

# This code chunk imports a dataset that contains mapping data for the boroughs of New York City. I wrangled
# the dataset by renaming all of the regions included in the dataset to just the name of the borough they are located in,
# so joining with future datasets would be simpler.


# map(database="nyc")
nyc <- map_data("nyc")
nyc <- nyc %>%
  mutate(Borough = case_when(grepl("Manhattan", region) ~ "Manhattan",
                             grepl("Bronx", region) ~"Bronx",
                             grepl("Brook", region) ~ "Brooklyn",
                             grepl("Queens", region) ~ "Queens",
                             grepl("Staten Island", region) ~ "Staten Island"))
nyc$subregion <- NULL

# This is an example of the graph that this dataset produces.
# nyc_borough_plot <- nyc %>%
#   ggplot() + 
#   geom_map(map=nyc,
#            color = "gray",
#            mapping = aes(x=long, 
#                y=lat, 
#                map_id=region)) +
#   coord_equal() +
#   theme_map()
```

```{r wrangling-historical-crime-1, message=FALSE, warning = FALSE}
# This code chunk wrangles historical felony data from 2000 to 2020 in New York City.
# importing raw historical felony data from csv
hist_felony_raw <- read_csv("data/seven-major-felony-offenses-2000-2020.csv")
hist_felony_data <- hist_felony_raw

# wrangling of first dataset. variables names are changed to make data more concise. data is pivoted longer
# to allow data to be sorted by year. The row that provides the total sum of felonies is removed as it is not needed.
hist_felony_data <- hist_felony_data  %>%
  rename(Felony = OFFENSE) %>%
  mutate(Felony = str_replace(Felony,
                              "MURDER & NON-NEGL. MANSLAUGHTER",
                              "MURDER"),
         Felony = str_replace(Felony,
                              "FELONY ASSAULT",
                              "ASSAULT"),
         Felony = str_replace(Felony,
                              "GRAND LARCENY OF MOTOR VEHICLE",
                              "GRAND LARCENY OF VEHICLE")) %>%
  pivot_longer(cols = `2000`:`2020`, 
               names_to = "Year", 
               values_to = "Number") %>%
  filter(Felony != "TOTAL SEVEN MAJOR FELONY OFFENSES")

```

```{r wrangling-historical-pop-1, message=FALSE, warning = FALSE}
# This code chunk wrangles historical population data for New York City.
# importing raw historical population data from csv
hist_borough_pop_raw <- read_csv("data/new-york-city-population-by-borough-1950-2040.csv")
hist_borough_pop <- hist_borough_pop_raw 

# wrangling of second dataset. the populations for the years 2000, 2010, and 2020 were selected. The total
# New York City population was selected for each year. The dataset was pivoted longer to make the data
# more usable. The individual borough populations were removed.
hist_borough_pop_tot <- hist_borough_pop %>%
  select(`Borough`,`2000`,`2010`,`2020`) %>%
  filter(Borough == "NYC Total") %>%
    pivot_longer(cols = `2000`:`2020`, 
               names_to = "Year", 
               values_to = "NycPop")
hist_borough_pop_tot$Borough = NULL

# the same raw population dataset is wrangled to display the populations for each borough. 
hist_borough_pop <- hist_borough_pop %>%
  select(`Borough`,`2000`,`2010`,`2020`) %>%
  filter(Borough != "NYC Total") 
hist_borough_pop_longer <- hist_borough_pop %>%
  pivot_longer(cols = `2000`:`2020`, 
               names_to = "Year", 
               values_to = "Pop")

# data for the year 2000 is selected from the dataset
nyc_pop_2000 <- hist_borough_pop %>%
  select(`Borough`,`2000`) %>%
  rename(Pop = `2000`)

# data for the year 2010 is selected from the dataset
nyc_pop_2010 <- hist_borough_pop %>%
  select(`Borough`,`2010`) %>%
  rename(Pop = `2010`)

# data for the year 2020 is selected from the dataset
nyc_pop_2020 <- hist_borough_pop %>%
  select(`Borough`,`2020`) %>%
  rename(Pop = `2020`)
```

```{r line-graph-1, message=FALSE, warning = FALSE}
# This code chunk generates a line graph that displays the evolution of the numbers of felonies committed from 2000 to 2020
# in New York City. This is total felonies, not per capita.

# Line graph showing all felonies in nyc from 2000-2020
colors = c(wes_palette("GrandBudapest1"), wes_palette("GrandBudapest2"))
hist_felony_plot_2 <- hist_felony_data %>%
  ggplot(aes(x = Year, y = Number, group = Felony, color = Felony)) +
  geom_line(size = 1) +
  labs(x = NULL, y = "Number of Cases", title = "Number of Felonies Committed", subtitle = "Over The Past Twenty Years in NYC") +
  scale_color_manual(values = colors) +
  theme(legend.position = "top", 
        axis.text.x = element_text(angle = 90),
        legend.title = element_blank())
hist_felony_plot_2
```

```{r wrangling-historical-crime-2, message=FALSE, warning = FALSE}
# This code chunk wrangles historical felony data by precinct, which allows the historical felony dataset to be connected to the New York City Borough mapping data.

# importing raw historical felony by precinct data from csv
hist_felony_by_precinct_raw <- read_csv("data/seven-major-felony-offenses-by-precinct-2000-2020.csv")
hist_felony_by_precinct_data <- hist_felony_by_precinct_raw 

# wrangling fourth dataset.
# dataset wrangled with FOUR pipes.
# column names are renamed for clarity, felony names are renamed for readability and simplicity, the dataset is pivoted longer to allow for sorting by year, total sum of felonies is removed because it is not needed, precinct numbers are used to assign rows to boroughs, the precinct column is removed because it is no longer needed, and the dataset is finally grouped by borough and felony to facilitate the next graphs.
hist_felony_by_precinct_data <- hist_felony_by_precinct_data %>%
  rename(Felony = CRIME,
         Precinct = PCT) %>%
  mutate(Felony = str_replace(Felony,
                              "MURDER & NON NEGL. MANSLAUGHTER",
                              "MURDER"),
         Felony = str_replace(Felony,
                              "FELONY ASSAULT",
                              "ASSAULT"),
         Felony = str_replace(Felony,
                              "GRAND LARCENY OF MOTOR VEHICLE",
                              "GRAND LARCENY OF VEHICLE")) %>%
  pivot_longer(cols = `2000`:`2020`, 
               names_to = "Year", 
               values_to = "Number") %>%
  filter(Felony != "TOTAL SEVEN MAJOR FELONY OFFENSES") %>%
  mutate(Borough = case_when((Precinct < 35) ~ "Manhattan",
                             (Precinct > 34 & Precinct < 53) ~ "Bronx",
                             (Precinct > 52 & Precinct < 94) ~ "Brooklyn",
                             (Precinct > 93 & Precinct < 120) ~ "Queens",
                             (Precinct > 119) ~ "Staten Island")) %>%
  select(-Precinct) %>%
  group_by(Borough, Felony)

# shows the total felonies of each type committed in each borough from 2000-2020. gets rid of na values in the number column.
hist_felony_by_precinct_data_tot <- hist_felony_by_precinct_data
hist_felony_by_precinct_data_tot <- hist_felony_by_precinct_data_tot %>%
  filter(!is.na(Number)) %>%
  ungroup() %>%
  group_by(Borough, Felony, Year) %>%
  mutate(Total = sum(Number)) %>%
  select(-Number) %>%
  distinct()
```

```{r wrangling-crime-per-capita, messafe=FALSE, warning=FALSE}
# This code chunk wrangles data to allow for the generation of graphs that display information regarding the total number of felonies committed in each borough per capita for the years 2000, 2010, and 2020.

# wrangling data to graph felonies committed by borough for 2000
felonies_boroughs_2000 <- hist_felony_by_precinct_data 
felonies_boroughs_2000 <- felonies_boroughs_2000 %>%
  filter(Year == "2000" & !is.na(Number)) %>%
  mutate(Total = sum(Number)) %>%
  select(-Number) %>%
  distinct() 

# shows the top felonies committed in each borough in 2000
felonies_boroughs_2000_top <- felonies_boroughs_2000 
felonies_boroughs_2000_top <- felonies_boroughs_2000_top %>%
  group_by(Borough) %>%
  top_n(1, Total)

# adds mapping data and population data
felonies_boroughs_2000_p <- felonies_boroughs_2000 %>%
  left_join(nyc_pop_2000, by = "Borough") %>%
  mutate(PerCapita = Total/Pop) %>%
  left_join(nyc, by = "Borough")

# ---

# wrangling data to graph felonies committed by borough for 2010
felonies_boroughs_2010 <- hist_felony_by_precinct_data 
felonies_boroughs_2010 <- felonies_boroughs_2010 %>%
  filter(Year == "2010" & !is.na(Number)) %>%
  mutate(Total = sum(Number)) %>%
  select(-Number) %>%
  distinct()

# shows the top felonies committed in each borough in 2010
felonies_boroughs_2010_top <- felonies_boroughs_2010 
felonies_boroughs_2010_top <- felonies_boroughs_2010_top %>%
  group_by(Borough) %>%
  top_n(1, Total)

# adds mapping data and population data
felonies_boroughs_2010_p <- felonies_boroughs_2010 %>%
  left_join(nyc_pop_2010, by = "Borough") %>%
  mutate(PerCapita = Total/Pop) %>%
  left_join(nyc, by = "Borough")

# ---

# wrangling data to graph felonies committed by borough for 2020
felonies_boroughs_2020 <- hist_felony_by_precinct_data 
felonies_boroughs_2020 <- felonies_boroughs_2020 %>%
  filter(Year == "2020" & !is.na(Number)) %>%
  mutate(Total = sum(Number)) %>%
  select(-Number) %>%
  distinct() 

# shows the top felonies committed in each borough in 2020
felonies_boroughs_2020_top <- felonies_boroughs_2020 
felonies_boroughs_2020_top <- felonies_boroughs_2020_top %>%
  group_by(Borough) %>%
  top_n(1, Total)

# adds mapping data and population data
felonies_boroughs_2020_p <- felonies_boroughs_2020 %>%
  left_join(nyc_pop_2020, by = "Borough") %>%
  mutate(PerCapita = Total/Pop) %>%
  left_join(nyc, by = "Borough")
```

```{r three-decade-borough-maps, message=FALSE, warning = FALSE}
# This code chunk generates three plots. Each plot shows the numbers of total felonies committed per capita in each borough of New York City in the years 2000, 2010, and 2020.
# Map of Number of Felonies Per Capita committed in each Borough in 2000.
felonies_boroughs_2000_plot <- felonies_boroughs_2000_p %>%
  mutate(Felony = str_replace(Felony,
                              "OF VEHICLE",
                              "AUTO")) %>%
  ggplot() +
  geom_map(map = nyc,
           color = "gray",
           mapping = aes(x = long,
               y = lat,
               map_id = region,
               fill = PerCapita)) +
  coord_equal() +
  theme_map() +
  theme(legend.key.width = unit(1.5, 'cm'),
        legend.key.height = unit(1, 'cm')) +
  scale_fill_distiller(palette="Blues", direction = 1) +
  facet_wrap(~Felony, labeller = labeller(group = label_wrap_gen(width = 50))) +
  labs(title = "Total Committed Felonies Per Capita", subtitle = "in 2000")

# Map of Number of Felonies committed in each Borough in 2010.
felonies_boroughs_2010_plot <- felonies_boroughs_2010_p %>%
  mutate(Felony = str_replace(Felony,
                              "OF VEHICLE",
                              "AUTO")) %>%
  ggplot() +
  geom_map(map = nyc,
           color = "gray",
           mapping = aes(x = long,
               y = lat,
               map_id = region,
               fill = PerCapita)) +
  coord_equal() +
  theme_map() +
  theme(legend.key.width = unit(1.5, 'cm'),
        legend.key.height = unit(1, 'cm')) +
  scale_fill_distiller(palette="Blues", direction = 1) +
  facet_wrap(~Felony) +
  labs(title = "Total Committed Felonies Per Capita", subtitle = "in 2010")

# Map of Number of Felonies committed in each Borough in 2020.
felonies_boroughs_2020_plot <- felonies_boroughs_2020_p %>%
  mutate(Felony = str_replace(Felony,
                              "OF VEHICLE",
                              "AUTO")) %>%
  ggplot() +
  geom_map(map = nyc,
           color = "gray",
           mapping = aes(x = long,
               y = lat,
               map_id = region,
               fill = PerCapita)) +
  coord_equal() +
  theme_map() +
  theme(legend.key.width = unit(1.5, 'cm'),
        legend.key.height = unit(1, 'cm')) +
  scale_fill_distiller(palette="Blues", direction = 1) +
  facet_wrap(~Felony) +
  labs(title = "Total Committed Felonies Per Capita", subtitle = "in 2020")

felonies_boroughs_2000_plot
felonies_boroughs_2010_plot
felonies_boroughs_2020_plot
```


```{r three-decade-bar-charts, message=FALSE, warning = FALSE}
# Three bar plots for felonies committed in nyc in 2000, 2010, and 2020.
colors = c(wes_palette("Darjeeling1"), wes_palette("Royal1"))
hist_felony_plot_3 <- hist_felony_data %>%
  filter(Year == "2000" |
         Year == "2010" |
         Year == "2020") %>%
  arrange(desc(Number)) %>%
  ggplot(aes(x = Felony,
           y = Number,
           fill = Felony)) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "Number", title = "Felonies Committed in Total in NYC in 2000, 2010, and 2020", fill = "") +
  scale_fill_manual(values = colors) +
  theme_minimal() +
  theme(legend.position = "top",
        plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank()) +
  facet_wrap(~Year)
hist_felony_plot_3

colors2 = c(wes_palette("GrandBudapest1"), wes_palette("GrandBudapest2"))
hist_felony_plot_4 <- hist_felony_data %>%
  filter(Year == "2000" |
         Year == "2010" |
         Year == "2020") %>%
  arrange(desc(Number)) %>%
  left_join(hist_borough_pop_tot, by="Year") %>%
  mutate(PerCap = Number/NycPop) %>%
  ggplot(aes(x = Felony,
           y = PerCap,
           fill = Felony)) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "Felonies Per Capita", title = "Felonies Committed Per Capita in NYC in 2000, 2010, and 2020", fill = "") +
  scale_fill_manual(values = colors2) +
  theme_minimal() +
  theme(legend.position = "top",
        plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank()) +
  facet_wrap(~Year)
hist_felony_plot_4
```

<br>
<br>
<br>
<br>
<br>

### Analysis

1. What were the most common felonies committed in the city of New York in the years 2000, 2010, and 2020?<br>
  + The most common felonies committed in New York City both by total amount and per capita were Grand Larceny in all three years. As seen in the graphs, Grand Larceny is far more common than other crimes. In New York City, Grand larceny involves the theft of property valued at over $1000. It also applies to the theft of any firearm or vehicle, and can apply to extortion and blackmail.<br>
<br>

```{r answer-1, echo=FALSE}
hist_felony_plot_3
hist_felony_plot_4
```

<br>
2. What were the most common felonies committed in each borough of New York in 2000, 2010, and 2020?<br>
  + The most common felonies committed in each borough are displayed in the tables below:
  In the year 2000 there were a variety of felonies being committed most often. As time goes on, the trends change. As seen across this report, as overall crime has decreased in New York, Grand Larceny has decreased the slowest, which has made it the most commonly committed felony in New York City today.
<br>

```{r answer-2, echo=FALSE}
# using kableExtra tables to display top felonies
felonies_boroughs_2000_top <- felonies_boroughs_2000_top[,c(2,3,1,4)]
felonies_boroughs_2000_top %>%
  select(-`Year`) %>%
  kbl(caption = "Most Commonly Committed felonies for each borough of NYC in 2000") %>%
  kable_classic(full_width = F, html_font = "Cambria", position = "center")

felonies_boroughs_2010_top <- felonies_boroughs_2010_top[,c(2,3,1,4)]
felonies_boroughs_2010_top %>%
  select(-`Year`) %>%
  kbl(caption = "Most Commonly Committed felonies for each borough of NYC in 2010") %>%
  kable_classic(full_width = F, html_font = "Cambria", position = "center")

felonies_boroughs_2020_top <- felonies_boroughs_2020_top[,c(2,3,1,4)]
felonies_boroughs_2020_top %>%
  select(-`Year`) %>%
  kbl(caption = "Most Commonly Committed felonies for each borough of NYC in 2020") %>%
  kable_classic(full_width = F, html_font = "Cambria", position = "center")
```

<br>
3. How have felony rates changed over the past two decades?<br>
  + Overall crime in New York has decreased. Both the overall amount of felonies and the amount of felonies per capita have decreased notably. In 2019, burglary and grand larceny of vehicles increased, while general grand larceny decreased. Also, there was a period of time around 2010 where grand larceny had decreased significantly. Overall, however, Grand Larceny has become the most commonly committed crime in New York City.<br><br>
<br> 

```{r answer-3, echo=FALSE}
hist_felony_plot_2
```

```{r answer-3-a, echo=FALSE}
felonies_boroughs_2000_plot
felonies_boroughs_2010_plot
felonies_boroughs_2020_plot
```


